Mode and tempo of promoter evolution as measured in the \textit{Paramecium aurelia} species complex

### 2019-03-12 manuscript.tex

\keyword{cis-regulatory regions, duplicate genes, paralogous genes, Paramecium, promoter evolution, TSS profiling.}

@@

Investigation across diverse eukaryotes has drawn attention to the important role of \textit{cis}-regulatory sequences in the evolution of gene expression \citep{Wittkopp:2008ki, Wittkopp:2011bc, Siepel:2014hd}. However, the \textit{cis}-regulatory sequence differences that accompany---or indeed underpin---species divergence remain largely unknown. One plausible explanation for this is the difficulty in predicting \textit{cis}-regulatory regions from genomic sequence alone. Accurate estimation of promoter positions requires empirical, usually functional genomic evidence. At present, the best high-throughput approach to identify promoters at genome-scale is Transcription Start Site (TSS) profiling, which includes CAGE \citep{Murata:gf} and RAMPAGE \citep{Batut:2013fu}, among others. While TSS profiling methods differ among themselves methadologically, these methods capture the 5$`$-ends of capped mRNAs, sequence their corresponding cDNAs and align the reads to the genome to identify the TSSs present within a given transcriptome. Clustering gene-adjacent TSSs defines a promoter, transcription start region (TSR), at single base-pair resolution, thereby providing genomic locations for \textit{cis}-regulatory regions within a species \citep{Lenhard:2012ena,Rach:2009ct}. TSS profling studies in a variety of model organisms has demonstrated that the shape of TSS distributions (\textit{i.e.} promoter shape) at promoters is related to the promoter class and correlates with the function of the associated gene \citep{Carninci:2006in, Rach:2009ct,Hoskins:2011io,Raborn:2016cr}. In addition, recent work indicates that promoter shape is itself a quantitative trait \citep{Schor:2017fw}, which has important implications for the understanding of the evolution of \textit{cis}-regulatory regions.      \\

>>>>

\keyword{\textit{cis}-regulatory regions, duplicate genes, paralogous genes, \textit{Paramecium}, promoter evolution, TSS profiling.}

@@

Investigations across diverse eukaryotes have drawn attention to the important role of \textit{cis}-regulatory sequences in the evolution of gene expression \citep{Wittkopp:2008ki, Wittkopp:2011bc, Siepel:2014hd}. Work over the previous two decades have emphasized the contribution of \textit{cis}-regulatory sequence differences toward changes in gene expression (insert ref to Wittkopp 2004).  However, the \textit{cis}-regulatory sequence differences that accompany---or indeed underpin---species divergence remain largely obscure. One plausible explanation for this is the difficulty in predicting \textit{cis}-regulatory regions from genomic sequence alone. Accurate estimation of promoter positions requires empirical, usually functional genomic evidence. At present, the best high-throughput approach to identify promoters at genome-scale is Transcription Start Site (TSS) profiling, which includes CAGE \citep{Murata:2014gpb} and RAMPAGE \citep{Batut:2013fu}, among others. While TSS profiling methods differ in technical between themselves, these protocols all capture the 5$`$-ends of capped mRNAs, sequence their corresponding cDNAs and align the reads to the genome to identify the TSSs present within a given transcriptome. Clustering gene-adjacent TSSs defines a promoter, transcription start region (TSR), at single base-pair resolution, thereby providing genomic locations for \textit{cis}-regulatory regions within a species \citep{Lenhard:2012ena,Rach:2009ct}. TSS profling studies in a variety of model organisms has demonstrated that the shape of TSS distributions (\textit{i.e.} promoter shape) at promoters is related to the promoter class and correlates with the function of the associated gene \citep{Carninci:2006in, Rach:2009ct,Hoskins:2011io,Raborn:2016cr}.  In addition, recent work done in \textit{Drosophila melanogaster} indicates that promoter shape is itself a quantitative trait \citep{Schor:2017fw}, a finding that carries valuable implications for the understanding of the evolution of \textit{cis}-regulatory regions. \\
